### 2016-03-15 chapters/background.tex
rework with regard to our advisors feedback

\subsection{Product configuration}
Through configuration a concrete product can be derived from a product line. Each valid configuration represents a specific \textit{variant} of the possible products.

During configuration a user selects or unselects features to his needs. This process requires a lot of domain knowledge on the one hand and detailed information about each single feature on the other. With growing numbers of possible features (and thus growing numbers of possible variants) configuration get more and more complex and turns out to be not trivial.

One of the purposes of the feature-oriented approach is saving the effort of creating a whole new product and deriving that product from a product line instead. The sheer overhead of configuration might even negate that benefit if configuration grows too complex.

\subsection{Constraints, contradictions, SAT-solver}
As stated above a feature model may contain constraints in the form of boolean expressions. Also, the feature-tree can be expressed as a boolean statement. The model shown in figure~\ref{img-fm} can be expressed as follows:
\begin{equation}
\begin{split}
	GraphLibrary\ \wedge\ Edges\\
	\wedge\ ((Directed\ \wedge \neg Undirected)\\
	\vee (\neg Directed\ \wedge\ Undirected))\\
	\wedge\ (Algorithms\ \Rightarrow\ (Number\ \vee\ Cycle))\\
	\wedge\ (Cycle\ \Rightarrow\ Directed)
\end{split}
\end{equation}\\
This formalism allows a configuration to be checked for validity. To do so each selected Feature is appended with a logical \textit{AND} and each specifically unselected feature is also appended with a logical \textit{AND} but gets negated. The resulting expression is then evaluated by a SAT-solver to check for satisfiability. Even during configuration this process can be applied to check for invalid partial configurations after each decision.

>>>>

\subsection{Product configuration}
The whole product line is represented by it's feature model, which itself is a set of features with specific interrelations. The process of configuration of a product line describes the steps to derive a product from the product line. To archive this, one has to select a subset of all the possible features within the product line to meet one's requirements. But not all subsets of features result in an actual product. The interrelations of the features restrict the possible combinations of features resulting in a \textit{valid} configuration. If only one of the requirements from the interrelations between the features is not met, no product can be created and the configuration as such is considered \textit{invalid}.

During configuration a user selects or unselects features to his needs. This process requires a lot of domain knowledge on the one hand and detailed information about each single feature on the other. With growing numbers of possible features (and thus growing numbers of possible products) configuration requires more and more effort. The sheer amount of time needed to iterate over all the features grows to be significant. But the even greater problem arises from the possible interactions and interrelations of features, which one has to keep track of during configuration. The overhead of configuration might even outright negate the benefit gained from using a product line instead of multiple stand-alone products.

{\color{red}TODO: introduce partial configuration}

\subsection{Constraints, contradictions, SAT-solver}
A feature-tree (from a feature model) can be expressed as a boolean statement. The mapping between the individual groups (sub-trees) and the corresponding boolean expression is shown in the folloing list:

{\color{red}TODO: Rules for
\begin{itemize}
	\item Or
	\item Alternative
	\item And
\end{itemize}
}
These particular groups are joined via a boolean \textit{AND}. For the example feature tree shown in figure~\ref{img-fm} the corresponding boolean expression is as follows:
\begin{equation}
\begin{split}
	GraphLibrary\ \wedge\ Edges\\
	\wedge\ ((Directed\ \wedge \neg Undirected)\\
	\vee (\neg Directed\ \wedge\ Undirected))
\end{split}
\end{equation}\\

As the feature model not only contains the feature tree, but also additional constraints, these constraints have to be considered in the boolean expression representing the complete feature model. They can be linked to one another via boolean \textit{AND}s. The connection to the previously created boolean expression of the feature tree is also made through a logical \textit{AND}. Thus, the complete resulting boolean expression for the feature model shown in figure~\ref{img-fm} is:
\begin{equation}
\begin{split}
	GraphLibrary\ \wedge\ Edges\\
	\wedge\ ((Directed\ \wedge \neg Undirected)\\
	\vee (\neg Directed\ \wedge\ Undirected))\\
	\wedge\ (Algorithms\ \Rightarrow\ (Number\ \vee\ Cycle))\\
	\wedge\ (Cycle\ \Rightarrow\ Directed)
\end{split}
\end{equation}\\
This formalism allows a configuration to be checked for validity. To do so each selected Feature is appended with a logical \textit{AND} and each specifically unselected feature is also appended with a logical \textit{AND} but gets negated. The resulting expression is then evaluated by a SAT-solver to check for satisfiability. Even during configuration this process can be applied to check for invalid partial configurations after each decision.

### 2016-03-15 chapters/concept.tex
rework with regard to our advisors feedback

Figure~\ref{img-workflow} shows the general process aimed to archive within this work. At the beginning there is only a given product line with the modular code artifacts in their specific subdirectories of the project. These code artifacts follow certain conventions to describe their integration in the context of the whole project.

Using these conventions a feature model gets extracted in which the hierarchies and dependencies of the code artifacts are included. This allows for much better overview or automated optimization, for example scanning for dead features via SAT-solver. The extracted feature model will also be used in the following step of configuration.

Configuration of product lines generally requires a good understanding of the given problems and the possible solutions to these problems as well as the specific implementations of these solutions. This domain knowledge is the most challenging part of configurations and restricts most of the possible users to do the configuration on their own.

To transfer the domain knowledge of experts to the product line a questionnaire is introduced. In the progress of implementing the product line experts also design a questionnaire in such a way that a possible user has to answer a given amount of questions to perform the configuration of a variant meeting his personal needs without him having to know all the details of the implementation of the product line or even the individual features. The two major steps of extraction and the creation and usage of the questionnaire are described in detail on the following sections.

\subsection{Extraction of a feature model}
To automatically extract a feature model out of a given software project the structure of that project must be algorithmically processable. To archive this, conventions are used for the project of which an overview follows:

\begin{itemize}
	\item All features are stored within a specified directory
	\item Each feature is stored in a separate directory
	\item A feature's parent features are stated in the name of the corresponding directory
	\begin{itemize}
		\item The complete hierarchy is displayed, except for the root feature
		\item Individual parent features are delimited with an unique symbol
	\end{itemize}
	\item Each feature directory contains a configuration file with a specified name containing:
	\begin{itemize}
		\item A descriptive name of the feature
		\item A description text for the feature
		\item All dependencies including the parent features
	\end{itemize}
\end{itemize}

All of these information are included in the resulting feature model. The most obvious correspondence lies within the hierarchy. The parent names are parsed from the directory name and -if existing- looked up in the existing partial feature model from the previously parsed features. As the directories are parsed in alphabetical order a parent feature will always be processed before it's child features.

After placing it in the correct position within the existing partial feature model each feature's configuration file gets processed. Each feature gets enriched with it's descriptive details and it's dependencies. The list of dependencies firstly gets reduced by the parent features as the feature models hierarchy already implements this kind of dependencies. The remaining dependencies are stated as constraints. To shorten the list of constraints shared dependencies between multiple features are combined to a single constraint.\\

@@

@@

\begin{lstlisting}
<configurationSurvey>
	<projectName>Name</projectName>
	<section id="0">
		<name>Section Name</name>
		<description>Section description</description>
	</section>
	<page id="0" sectionId="0">
		<question>Question for the user</question>
		<answers type="alternative">
			<answer nextPageId="1">
				<label>Answer label</label>
				<description>Answer description</description>
				<dependencies>
					<feature selection="false">Unselected feature</feature>
					<feature>selected feature</feature>
				</dependencies>
			</answer>
		</answers>
	</page>
</configurationSurvey>
\end{lstlisting}

\begin{tabular}{ l | l | l }
\textbf{Tag} & \textbf{attributes} & \textbf{child tags}\\
\hline
\hline
configurationSurvey & & projectName, section, page\\
\hline
section & id & name, description\\
\hline
page & id, sectionId & question, answers\\
\hline
answers & type & answer\\
\hline
answer & nextPageId & label, description, dependencies\\
\hline
dependencies & & feature\\
\hline
feature & selection & \\
\end{tabular}\\

>>>>

This work was build around the usage for Odoo\footnote{https://www.odoo.com/}, an open-source tool for enterprise-resource-planing written mainly in python\footnote{https://www.python.org/}. In this chapter, we will explain the workflow shown in figure~\ref{img-workflow} on the basis of the experiences we made with Odoo.

Odoo itself is programmed as a product line. It has a lot of modular code artifacts in their specific subdirectories of the project. These code artifacts follow certain conventions to describe their integration in the context of the whole project.

Using these conventions we were able to extract a feature model in which the hierarchies and dependencies of the code artifacts are included. This allows for much better overview over the product line and it's capabilities. During our efforts to implement a automated generation of a feature model we encountered some problems for which we had to find solutions/workarounds.

@@

The extracted feature model will also be used in the following step of configuration.

Configuration of product lines generally requires a good understanding of the given problems and the possible solutions to these problems as well as the specific implementations of these solutions. This domain knowledge is the most challenging part of configurations and restricts most of the possible users to do the configuration on their own. The same goes of course for Odoo and the various components contained in the project.

To empower possible users to configure Odoo on their own, we had the idea to use a questionnaire. In the progress of implementation, experts also design a questionnaire in such a way that a possible user has to answer a given amount of questions to perform the configuration of a product meeting his personal needs without him having to know all the details of the implementation of the product line or even the individual features. This effectively redesigns the process of configuration in such a way, that a user is independent of the knowledge about the details of the implementation and can focus on tailoring the product line to his specific use-case.

The two major steps of extraction of the feature model and the creation and usage of the questionnaire are described in detail on the following sections.

\subsection{Extraction of a feature model}
To automatically extract a feature model out of a given software project the structure of that project must be algorithmically processable. We had a thorough look at the source code of Odoo and found it to be structured according to the following rules:

\begin{itemize}
	\item All features are stored within a specified directory
	\item Each feature is stored in a separate sub-directory
	\item A feature's parent features are stated in the name of the corresponding directory
	\begin{itemize}
		\item The complete hierarchy is displayed, except for the root feature
		\item Individual parent features are delimited with an unique symbol
	\end{itemize}
	\item Each feature directory contains a descriptive file with a specified name containing:
	\begin{itemize}
		\item A descriptive name of the feature
		\item A description text for the feature
		\item All dependencies including the parent features
	\end{itemize}
\end{itemize}

Following these rules we were able to automatically extract the information needed to generate a feature model. For this procedure we implemented a plugin for FeatureIDE which processes the extracted information to a feature model. The procedure to generate a feature model is stated following.

 The most obvious correspondence lies within the hierarchy. The parent names are parsed from the directory name and -if existing- looked up in the existing partial feature model from the previously parsed features. As the directories are parsed in alphabetical order and this way of ordering places e.g. ``Feature1'' in front of ``Feature1\_Feature'', a parent feature will always be processed before it's child features.

After placing it in the correct position within the existing partial feature model each feature's descriptive file gets processed. Each feature gets enriched with it's descriptive details and it's dependencies. The list of dependencies firstly gets reduced by the parent features as the feature models hierarchy already implements this kind of dependencies. The remaining dependencies are stated as constraints. To shorten the list of constraints shared dependencies between multiple features are combined to a single constraint.

@@

\begin{lstlisting}
<configurationSurvey>
	<projectName>Name</projectName>
	<section id="0">
		<name>Section Name</name>
		<description>Section description</description>
	</section>
	<page id="0" sectionId="0">
		<question>Question for the user</question>
		<answers type="alternative">
			<answer nextPageId="1">
				<label>Answer label</label>
				<description>
					Answer description
				</description>
				<dependencies>
					<feature selection="false">
						Unselected feature
					</feature>
					<feature>selected feature</feature>
				</dependencies>
			</answer>
		</answers>
	</page>
</configurationSurvey>
\end{lstlisting}

\begin{tabulary}{\linewidth}{LLL}
\textbf{Tag} & \textbf{attributes} & \textbf{child tags}\\
\hline
configurationSurvey & & projectName, section, page\\
section & id & name, description\\
page & id, sectionId & question, answers\\
answers & type & answer\\
answer & nextPageId & label, description, dependencies\\
dependencies & & feature\\
feature & selection & \\
\end{tabulary}\vspace{2.5em}
